Improving \textsl{Gaia} parallaxes with a data-driven model \\ of the color--magnitude diagram

### 2017-05-30 paper/paper.tex
fixed all XXXs and YYYs

Conversion of a noisy parallax measurement into a posterior belief over distances requires inference with a prior.
Usually this prior represents some kind of belief about the Milky Way.
However, there is multi-band photometry for a large fraction of the \gaia\ \tgas\ catalog from imaging surveys;
this imaging is incredibly informative about stellar distances.
Here we use color information of $1.2$ million \tgas\ stars from \tmass\ to build a noise-deconvolved empirical prior distribution for stars in color--magnitude space.
This data-driven model contains no knowledge of the physics of stellar interiors or photospheres, nor of the Milky Way, but rather derives its precision from its generative model of noisy parallax measurements and an assumption of stationarity.
We use the Extreme Deconvolution (\xd) algorithm, which is an Empirical Bayes approximation to a full hierarchical model of the true parallax and photometry of every star.
The algorithm is run not in absolute-magnitude space but in a transformed space where the measurement uncertainty is closer to Gaussian.
The \xd-optimized prior is used to perform parallax inferences for every star, yielding a precise photometric parallax estimate and uncertainty (and full posterior) for every star.
Our posterior photometric parallax estimates are more precise than the \tgas\ catalog entries by a median factor of 1.2, and a 75, 90, and 99 percentile factor of 1.6, 2.3, and 4.9, respectively; and more precise than previous examples of Bayesian distance estimates by a median factor of 1.2, and a 75, 90 and 99 percentile factor of 1.5, 1.6, and 2.1, respectively.
%The precision can be attributed to the statistics concept of shrinkage.
We independently validate our distances by looking at members of a Milky Way star cluster M67, which is not visible as a cluster in the \tgas\ parallax estimates, but appears clearly as a cluster in our posterior parallax estimates.
All our results, including a posterior parallax and distance sampling for [ZZZ] \tgas\ stars, are available in companion electronic tables.
\end{abstract}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{delta.png}
\caption{Fractional Change in Variance: The left panel shows the natural log of the fractional change in the variance of the posterior relative to the likelihood as a function of the stellar color. The coloring of the points in both panels is by the fractional change in variance to guide the coloring of the right panel. The blue lines are the $1\sigma$ and $2\sigma$ contours. The median fractional change is XXX. Some posteriors have significantly smaller variances shown in black, but some actually increase in variance, shown as the yellow bump at $(J-K)^C$ about XXX. The right panel shows the posterior point estimates on the CMD colored by fraction change in variance. There is significant changes for the red giant branch stars, as well as some main sequence stars.}
\label{fig:delta}
\end{figure}

@@

%How this compares to other Bayesian attempts at inferring distances,
%like, eg, the Tri--Coryn papers. Favorably! And with \emph{fewer}
%assumptions, in fact! What form should this comparison be in?
There are other options for prior information when inferring distances to stars. One option is the exponentially declining spatial density of stars (cite: Tri, CBJ). It is parameterized by a scale length, and has the nice property that it is smooth out to very large distances so the posteriors are also very smooth. It is a fairly weak prior. It sets a smooth scale distance beyond which the posterior struggles to maintain much probability, and also converts a negative parallax likelihood into a posterior of very small parallax. It is a fairly broad prior so the median \gaia\ star has an increased posterior variance of XXX. It has the largest effect on the most noisy and negative parallaxes, and has the nice property of bringing these noisy or negative measurements to reasonable parallaxes. Given this prior, what does the observed \cmd\ look like?  The \cmd\ is broad and the red clump highly stretched.

>>>>

Conversion of a noisy parallax measurement into a posterior belief over distances requires inference with a prior.
Usually this prior represents some kind of belief about the Milky Way.
However, there is multi-band photometry for a large fraction of the \gaia\ \tgas\ catalog from imaging surveys;
this imaging is incredibly informative about stellar distances.
Here we use color information of $1.2$ million \tgas\ stars from \tmass\ to build a noise-deconvolved empirical prior distribution for stars in color--magnitude space.
This data-driven model contains no knowledge of the physics of stellar interiors or photospheres, nor of the Milky Way, but rather derives its precision from its generative model of noisy parallax measurements and an assumption of stationarity.
We use the Extreme Deconvolution (\xd) algorithm, which is an Empirical Bayes approximation to a full hierarchical model of the true parallax and photometry of every star.
The algorithm is run not in absolute-magnitude space but in a transformed space where the measurement uncertainty is closer to Gaussian.
The \xd-optimized prior is used to perform parallax inferences for every star, yielding a precise photometric parallax estimate and uncertainty (and full posterior) for every star.
Our posterior photometric parallax estimates are more precise than the \tgas\ catalog entries by a median factor of 1.2, and a 75, 90, and 99 percentile factor of 1.6, 2.3, and 4.9, respectively; and more precise than previous examples of Bayesian distance estimates by a median factor of 1.2, and a 75, 90 and 99 percentile factor of 1.5, 1.6, and 2.1, respectively.
%The precision can be attributed to the statistics concept of shrinkage.
We independently validate our distances by looking at members of a Milky Way star cluster M67, which is not visible as a cluster in the \tgas\ parallax estimates, but appears clearly as a cluster in our posterior parallax estimates.
All our results, including a posterior parallax and distance sampling for 1.4 million \tgas\ stars, are available in companion electronic tables.
\end{abstract}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{delta.png}
\caption{Fractional Change in Variance: The left panel shows the natural log of the fractional change in the variance of the posterior relative to the likelihood as a function of the stellar color. The coloring of the points in both panels is by the fractional change in variance to guide the coloring of the right panel. The blue lines are the $1\sigma$ and $2\sigma$ contours. The median fractional change is 1.2. Some posteriors have significantly smaller variances shown in black, but some actually increase in variance, shown as the yellow bump at $(J-K)^C$ of about 0.6. The right panel shows the posterior point estimates on the CMD colored by fraction change in variance. There is significant changes for the red giant branch stars, as well as some main sequence stars.}
\label{fig:delta}
\end{figure}

@@

%How this compares to other Bayesian attempts at inferring distances,
%like, eg, the Tri--Coryn papers. Favorably! And with \emph{fewer}
%assumptions, in fact! What form should this comparison be in?
There are other options for prior information when inferring distances to stars. One option is the exponentially declining spatial density of stars (cite: Tri, CBJ). It is parameterized by a scale length, and has the nice property that it is smooth out to very large distances so the posteriors are also very smooth. It is a fairly weak prior. It sets a smooth scale distance beyond which the posterior struggles to maintain much probability, and also converts a negative parallax likelihood into a posterior of very small parallax. It is a fairly broad prior so the median \gaia\ star has a posterior variance equal to the likelihood variance. It has the largest effect on the most noisy and negative parallaxes, and has the nice property of bringing these noisy or negative measurements to reasonable parallaxes. Given this prior, what does the observed \cmd\ look like?  The \cmd\ is broad and the red clump highly stretched.
